---
title: "Stroke data EDA"
author: "Guy Maskall"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: pdf_document
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document downloads and performs an initial view of a 
[stroke dataset](https://www.kaggle.com/asaumya/healthcare-dataset-stroke-data).

```{r read}
stroke <- read_csv("train_2v.csv")
stroke %>% glimpse
```

## Initial data QA

Check id is unique and for any missing values.

### Uniqueness of ID

```{r data_quality_check}
stroke %>%
    count(id) %>%
    filter(n > 1)
stroke <- stroke %>% select(-id)
```

### Check for NA

Do any features have any NA?

```{r na_check}
# summarise_all soon to be superceded in dplyr 1.0.0
stroke %>% 
    summarise_all(~sum(is.na(.))) %>%
    t
```

We have some missing bmi and quite a lot of missing smoking_status.
Is there any obvious pattern in the missing values, or relationship
between the instances?

```{r na_patterns}
stroke %>% 
    mutate(bmi_na = if_else(is.na(bmi), "yes", "no")) %>% 
    filter(bmi_na == "yes" | is.na(smoking_status)) %>% 
    group_by(bmi_na) %>% 
    count(smoking_status)
```

From the above, probably the missing BMI values are missing at random,
certainly with respect to smoking,
seeing as they're broadly evenly distributed with smoking status.
The largest single group of missing values is that for smoking status
(where BMI is not missing).

## Stroke and smoking

We saw above that most of the missing values are in the smoking status
column. How does this seem to relate to stroke?

```{r smoking_stroke}
stroke %>% 
    count(smoking_status, stroke) %>% 
    pivot_wider(names_from=stroke, values_from=n) %>% 
    mutate(stroke_pc = 100*`1`/`0`)
```

From the above, we see that *smokes* and *never smoked* have a similar
rate of incidence of stroke, *formerly smoked* seems a bit higher,
and the *NA* smoking status group a bit lower. On the face of it, there
are some curious things here. How does it makes sense that "formerly
smoked" has a higher incidence of stroke than *smokes*, for example?
Could there be other factors involved, for example people forced to
give up smoking because of another health issue or perhaps taking
up unhealthy eating habits to compensate for no longer smoking?

Undersampling the majority
class repeatedly generates some consistent results wrt age and may
implicate smoking, depending on the composition of the groups.

### Modeling stroke risk using all available features

We balance the dataset, which means undersampling the
non-stroke group, and perform logistic regression.

```{r initial_glm}
set.seed(47)
complete_cases <- stroke %>% 
    filter(complete.cases(.))
run_glm <- function() {
    glm(stroke ~ ., family=gaussian, complete_cases %>% 
            group_by(stroke) %>%  
            sample_n(min(group_size(.)))) %>% 
    summary
}

print(run_glm())
print(run_glm())
print(run_glm())

```

We repeated the undersampling a couple of times above because we'd
noticed some variability in the results for different samples drawn
from the majority class. Age is pretty consistently strongly 
significant. This is no great surprise for stroke risk. Hypertension
and heart disease, and particularly average glucose level, are also
frequently implicated. The signs of their coefficients are all intuitive
as well; they are positive contributors to stroke risk.

We should pause here before continuing to dive into inference.
We don't know how the data were sampled and we have multiple
potentially confounding variables, especially categorical ones.
We should stress that we are still exploring potential relationships
amongst the variables, rather than attempting a rigorous statistical
inference. There is a hint that smoking may be a risk factor for
stroke.

We have some encouraging, and plausible, patterns, but
we should remember the words of Ronald Fisher:

> To consult the statistician after an experiment is finished is 
> often merely to ask him to conduct a post mortem examination. 
> He can perhaps say what the experiment died of.



Hypertension is binary, as is heart disease, and themselves
are imbalanced:

```{r hypertension_heart_imbalance}
stroke %>% count(hypertension)
stroke %>% count(heart_disease)
```

## Stroke and age

The relationship between stroke risk and age is clear in the
histograms below:

```{r age_stroke_hist}
stroke %>% 
    filter(gender != "Other") %>% 
    ggplot() + 
    geom_histogram(aes(x = age, fill = stroke==1)) + 
    facet_wrap(~gender) +
    labs(fill="Stroke")
```

We have more women than men, perhaps up to twice as many. The age
distribution for men is fairly flat, whereas for women there seem to
be more weight in the middle but fairly symmetric. The increasing
prevalence of stroke with age is clear.

## Age and smoking

Now we return to looking at smoking status. We suggested above that
there might be an occasional sign of a relationship between smoking
status and stroke, but we saw a very clear signal that stroke risk
was age related. Is there a relationship between age and smoking
status?

```{r age_smoking}
stroke %>% 
    ggplot() + 
    geom_density(aes(x = age, colour = smoking_status)) +
    labs(title="Age distribution with smoking status",
         subtitle="(full dataset)")
```

The distributions above contain much detail we can now put into
perspective. The missing values for smoking status are clearly
dominated by children. This makes total sense. It's quite
intuitive that the data pertaining to children did not include
a smoking status. This also means that simply omitting samples
missing a smoking status is fundamentally flawed because this is
to greatly bias the data away from younger people. If the desire
was to model stroke risk only for adults, that's one thing, but then
this should be explicitly done on age, not accidentally by smoking
status.

We can see another clear pattern as well. The category of ex-smokers
("formerly smoked") is heavily weighted to older ages. Again, this
is an intuitive result given a moment's thought. In order to be a
former smoker, one must first have been a smoker. A 20 year old has
had only a few potential smoking years, whereas a 60 year old has
many years of opportunity to be a smoker before giving up.

The young, below the early twenties, who have never smoked,
dominate those who do smoke. This effect alone would bias
the group who've never smoked to be younger, and so at less
age-related risk of stroke. Having said that, both the
*never smoked* and *formerly smoked* groups have an uptick at
the oldest ages that is lacking in the *smokes* group. This
could perhaps be a survival bias. It's not impossible that
the suppression of the oldest ages from the smoking group
might even make smoking appear protective of a stroke in some
cases; the truth, of course, would be closer to the fact that
smokers die of smoking-related disease before the stroke
risk ramps up.

In short, there seem some powerful interactions between age
and smoking status.

Can we further illustrate the linkage between age and
smoking status wrt stroke risk?

Firstly, with a bit of trial and error, we found a partitioning
of the data that suggested a mildly significant effect of smoking:

```{r age_smoke_link}
set.seed(100)
cc_balanced <- complete_cases %>% 
    group_by(stroke) %>% 
    sample_n(min(group_size(.)))
glm(stroke ~ age + smoking_status, 
    family=gaussian, cc_balanced) %>% 
    summary
```

So taking just age and smoking status (from the complete cases
data) we see a slight suggestion of a positive risk from being
a smoker compared to being a former smoker. Although the p-value
is quite high, the sign associated with having never smoked is
negative, which is in the intuitive direction, but the data really
can't be said to provide sufficient evidence for this.

```{r age_smoke_link_interaction}
set.seed(100)
cc_balanced <- complete_cases %>% 
    group_by(stroke) %>% 
    sample_n(min(group_size(.)))
glm(stroke ~ age * smoking_status, 
    family=gaussian, cc_balanced) %>% 
    summary
```

By now including interactions between age and smoking status,
the p-values associated with smoking are huge. Whilst we may
intuit that smoking could be a stroke risk, my gut feeling is
that in this dataset, smoking status is largely acting as a 
proxy for age. We can demonstrate this another way by modelling
age on smoking status:

```{r age_from_smoking_model}
set.seed(100)
glm(age ~ smoking_status,
    family=gaussian,
    complete_cases %>%
        group_by(smoking_status) %>%
        sample_n(min(group_size(.)))) %>%
summary
```

This shows us highly significant effects. On average, those
who smoke are 8 years younger than those who are former smokers,
and those who have never smoked tend to be younger still.
Given the powerful effect of age on stroke risk, we clearly have
the potential for smoking status to be a proxy (potentially
false) risk factor for stroke.


It is quite possible that smoking status has
some predictive power in addition to other features, but
we have to admit we don't really see any evidence of it
in this exploration. Given the quantity of samples missing
a smoking status, we should be far more comfortable dropping
this feature than dropping the samples with missing smoking
status.
